---
title: "NHSRDatasets meets runcharter"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(NHSRdatasets)
library(runcharter) # remotes::install_github("johnmackintosh/runcharter)
library(dplyr)
library(skimr)

```

## Background

The [NHSRDatasets](https://cran.r-project.org/package=NHSRdatasets) package made it to CRAN recently, and as it is designed for use by NHS data analysts, and I am an NHS data analyst, let's take a look at it.
Thanks to Chris Mainey and Tom Jemmett for getting this together.

## Load packages and data

As above let's load what we need for this session. 
The runcharter package is built using data.table, but I'm using dplyr in this main section to show that you don't need to know data.table to use it.

However- seriously, do take a look at data.table.
It's not has hard to understand as some might have you believe. A little bit of effort pays off.
You can load the package from github using the remotes package.
(I've managed to install it on Windows and Ubuntu. 
Mac user? No idea, I'm a Band 6  UK public sector data analyst, so can only dream of such luxuries).


```{r, echo= FALSE}
library(NHSRdatasets)
library(runcharter) ## remotes::install_github("johnmackintosh/runcharter")
library(dplyr)
```

```{r,echo=TRUE}
ae <- data("ae_attendances") # a promise
ae <- ae_attendances #  a string
rm(ae_attendances) # just tying 'ae' brings it to life in the environment
```


Let's have a look at the data:

```{r}
glimpse(ae)
```


Lot's of factors. I'm actually very grateful for this package, as it caused me major issues when I first tried to plot this data using an earlier version of runcharter. I hadn't considered factors as a possible grouping variable, which was a pretty big miss. 

There's way too much data for my tiny laptop screen, so I will filter the data for type 1 departments -  the package help gives us a [great link to explain what this means](https://www.datadictionary.nhs.uk/data_dictionary/attributes/a/acc/accident_and_emergency_department_type_de.asp)



```{r}
type1 <- ae %>% 
  filter(type == 1) %>% 
  arrange(period)

# plot attendances
p <- runcharter(type1,
                med_rows = 13, # median of first 13 points
                runlength = 9, # find a run of 9 consecutive points
                direction = "above", # find run above the original median
                datecol = "period",
                grpvar = "org_code",
                yval = "attendances")

```

The runcharter function returns both a plot, and a data.table/ data.frame showing a summary of any runs in the desired direction (I'm assuming folk reading this have a passing knowledge of run charts, but if not, you can look at the [package vignette](https://www.johnmackintosh.com/runcharter/articles/runcharter.html), which is the cause of most of my commits!!)

Don't try loading the plot right now, because it is huge, and takes ages. 
If we look at the summary dataframe,we can see 210 rows, a fairly decent portion of which relate to  significant increases above the original median value

```{r}
p$sustained
```


Let's use skimr to get a sense of this

```{r}
skimr::skim(p$sustained)
```


To keep this manageable, I'm going to filter out for areas that have median admissions > 10000 (based on the first 13 data points)

```{r}
high_admits <- p$sustained %>% 
  filter(median > 10000 & run_type == "sustained") %>%
  select(org_code)
```

Then I change the org_code from factor to character, and pull out unique values.
I'm sure there is a slicker way of doing this, but it's getting late, and I don't get paid for this..

I use the result to create a smaller data frame

```{r}
high_admits$org_code <- as.character(high_admits$org_code)


type1_high <- type1 %>% 
  filter(org_code %in% high_admits$org_code)
```

And now I can produce a plot that fits on screen. 
I've made the individual scales free along the y axis, and added titles etc

```{r runcharter-final}
p2 <- runcharter(type1_high,
                 med_rows = 13, # median of first 13 points as before
                 runlength = 9, # find a run of 9 consecutive points
                 direction = "above",
                 datecol = "period",
                 grpvar = "org_code",
                 yval = "attendances", 
                 facet_scales = "free_y",
                 facet_cols = 4,
                 chart_title = "Increased attendances in selected Type 1 AE depts",
                 chart_subtitle = "Data covers 2016/17 to 2018/19",
                 chart_caption = "Source : NHSRDatasets",
                 chart_breaks = "6 months")

```


Let's look at the sustained dataframe

```{r}
p2$sustained
```


And of course, the plot itself

```{r, fig.height=7, fig.width=9}
p2$runchart
```

I haven't looked into the actual data too much, but there are some interesting little facets here - what's the story with RDE, RRK and RTG for example?
I don't know which Trusts these codes represent, but they show a marked increase.

The RGN (top right) and RVV (mid left) show the reason why I worked on this package - we can see that there has been more than one increase. Performing this analysis in Excel is not much fun after a while. 

There is a lot more I can look at with this package, and we in the NHS-R community are always [happy to receive more datasets for inclusion](https://github.com/nhs-r-community/NHSRdatasets).


