---
title: "A tiny wee look at group_map and group_split in dplyr"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## GitHub Documents
Dplyr 0.8.0 launched recently.

Two new functions have been catching my eye :
group_map and group_split.

The aim of this post - take a first look at these and try and get a new blog post up in github
Time now: 23:38(UK)



## What it says on the tin

Load packages and create a function for demo purposes , basically grab the first 5 rows of whatever we throw at it:

```{r}
library(nycflights13)
library(dplyr)

data(flights)

header <- function(x,n = 5){
  head(x,n)
  }
```

Let's check the function works first of all:


```{r, echo=FALSE}
flights %>% 
  group_by(origin) %>% 
  header()
```

Good, that just returns 5 rows from the entire datset. 
Now let's replace group_by with group_split, and group by origin

```{r, echo = FALSE}
split_test <- flights %>% 
group_by(origin) 

```
Then we use group_split to create individual tibbles for each group

```{r}
group_split(split_test)

```

We can verify how the grouping was carried out by using group_keys:
```{r}


group_keys(split_test)
```



Now, let's get rid of that, and try group_map, again, by origin. 
We may as well use my highly pointless function.
If all goes well, we will return a tibble of 15 rows


```{r}
rm(split_test)
test5 <- flights %>% 
  group_by(origin) %>% 
 group_map(~ header(.x))
```


The dplyr help tells us that '.x' refers to the subset of rows for each group that is passed in. That returned 5 rows per origin, as expected.Let's try again and obtain some more rows per group:


```{r}

test10 <- flights %>% 
  group_by(origin) %>% 
 group_map(~ header(.x,10L))

```

Now we have 30 observations in our new tibble, so this all looks good.
Can we split and map together?

```{r}
#test_group_and_map <- flights %>% 
 # group_split(origin) %>% 
 #group_map(~ first5(.x,20))
```

No we cannot. 
Group_map works on grouped tibbles, and by splitting, we're passing it a list. 
Instant fail. RTFM folks.

At this point, I can see group_map being pretty useful, but I will need to do a bit more work on the use case for group_split - at present, I'm not seeing what it offers over and above group_by. 
As always, feel free to chip in if you have any thoughts or answers





